---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

File is unzipped.  
The data is read using read.csv  
The sqldf library is used to manipulate the data (throughout the code).  

```{r part_1_load_preprocess, echo=TRUE}
library(sqldf)
library(lattice)
unzip("activity.zip")
data <- read.csv("activity.csv")
data_per_day <- sqldf("select date,sum(steps) total_steps 
                      from data 
                      where steps is not null 
                      group by date")
```


## What is mean total number of steps taken per day?

```{r part_2_mean, echo=TRUE}
hist(data_per_day$total_steps,xlab="Total number of steps",main="Total number of steps taken each day", col="blue")
mean_val <- mean(data_per_day$total_steps)
mean_val
median_val <- median(data_per_day$total_steps)
median_val
```

The mean of total steps taken each day is `r format(mean_val,digits=7)`.  
The median of total steps taken each day is `r median_val`. 

## What is the average daily activity pattern?
1. Calculating for each interval - the average of steps taken, accross all days.
2. Plotting a time series of the intervals (x) and the avg of steps taken (y).
3. Finding the interval with the maximum average number of steps. 
```{r part_3_avg_daily, echo=TRUE}
avg_steps_per_interval <- sqldf("select interval, avg(steps) avg_steps 
                                from data  
                                group by interval")
plot(avg_steps_per_interval$interval, avg_steps_per_interval$avg_steps, type="l",
     xlab="Interval",ylab="avg steps",main="Avg steps taken per interval")

max_val <- sqldf("select interval, avg_steps from avg_steps_per_interval where avg_steps = (select max(avg_steps) from avg_steps_per_interval)")

max_val
```
The 5 minute interval with the maximum average steps taken is interval # `r max_val$interval`. The average steps taken in this interval is `r format(max_val$avg_steps,digits=5)`.    


## Imputing missing values
```{r part_4_nulls, echo=TRUE}
null_cnt <- sqldf("select count(*) from data where steps is null")
null_cnt
```

The number of 5-mintue intervals with NA values is `r null_cnt`  


The strategy for filling all NA values is the avg of steps for that 5-minute interval across all days.  
If a NA (null) value is detected, the avg_steps value of the same interval is used (from the calculations done in the previous section - part 3)

```{r part_4_imputing, echo=TRUE}
data_imputed <- sqldf("select 
                    case when steps is null then avg_steps else steps end i_steps , 
                    date, data.interval, avg_steps 
                  from data , avg_steps_per_interval 
                  where data.interval = avg_steps_per_interval.interval")


data_per_day_imputed <- 
  sqldf("select date,sum(i_steps) total_steps 
        from data_imputed 
        group by date")

hist(data_per_day_imputed$total_steps,xlab="Total number of steps",main="Total number of steps taken each day (data imputed)", col="red")
mean_i <- mean(data_per_day_imputed$total_steps)
mean_i
median_i <- median(data_per_day_imputed$total_steps)
median_i
```

The mean of total steps taken each day (after filling NA values) is `r format(mean_i,digits=7)`.  
The median of total steps taken each day (after filling NA values) is `r format(median_i,digits=7)`. 

The mean has not changed. 
The median has changed, and is now equal to the mean.

## Are there differences in activity patterns between weekdays and weekends?

Calculating the average steps for each interval - seperated to weekday and weekend intervals.  
Then, Plotting the time series of weekday days and weekend days.
```{r part_5_weekdays, echo=TRUE}
data_imputed$date_type <- sapply(weekdays(as.Date(data_imputed$date)),switch,
                                 Monday='weekday',
                                 Tuesday='weekday',
                                 Wednesday='weekday',
                                 Thursday='weekday',
                                 Friday='weekday',
                                 Saturday='weekend',
                                 Sunday='weekend')
data_imputed_avg <- sqldf("select interval, date_type, avg(i_steps) avg_steps
                          from data_imputed 
                          group by interval, date_type")
par(mfrow=c(2,1))

xyplot(data_imputed_avg$avg_steps~data_imputed_avg$interval | data_imputed_avg$date_type, type="l",layout=c(1,2),xlab="Interval",ylab="Number Of Steps")
```